Geometric variational problems

### 2020-11-05 NSF2020/Description/specificaim3_ver2.tex
working on numerics

The most significant challenges of a BIE formulation include solving
dense linear systems, developing preconditioning strategies, and
developing quadrature methods for integrands that are nearly singular.
These challenges are described in section~\ref{subsec:NumericalIssues}.
Then, in section~\ref{subsec:timeStepping}, time stepping strategies are
described, and stable BIE formulations for fluctuating hydrodynamics are
in section~\ref{subsec:fluctuating}.

@@

Discretizations of carefully chosen BIEs can be solved with a
mesh-independent number of GMRES
iterations~\cite{cam-ips-kel-mey-xue1996}. Therefore, the required CPU
time is proportional to the cost of a matrix-vector multiplication that
can be done in optimal or near-optimal time with the fast multiple
method (FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2, fmm3,
fmm4, fmm6, fmm7, fmm8}. PI Quaife is experienced with applying FMMs for
both the Stokes equation~\cite{qua-bir2014, bys-sha-qua2020} and the
screened Laplace equation~\cite{kro-qua2011, qua2011}. Alternatively,
fast direct solvers can be used to avoid any iteration~\cite{fds1, fds2,
fds3, fds4, fds5, fds6, fds7, fds8, ho2016cpam2, ho2016cpam1,
minden2016, minden2017siammms}, but direct solvers often require a large
amount of initial overhead, and this makes them less useful for problems
with moving geometries.  Alternatively, techniques in fast direct
solvers can be used to develop efficient preconditioning strategies. PI
Quaife used the inverse Fast Multipole Method
(IFMM)~\cite{cou-pou-dar2017, qua-cou-dar2018} to precondition Stokes
equations in a porous media, and a suite of preconditioners ranging from
block-diagonal to the IFMM will be considered.

Nearly touching bodies is ubiquitous in self-assembly of amphiphilic
particles, and such near-contact results in nearly-singular integrands
in both two and three dimensions. Quadrature methods to address such
integrands has received a lot of attention in two and three
dimensions~\cite{alpert, kapur, sidi, duffy, bruno1, bruno2, davis_1984,
graglia_2008, hackbusch_sauter_1994, jarvenpaa_2003, khayat_2005,
schwab_1992, ying_2006, beale1, beale2, goodman_1990, haroldson_1998,
lowengrub_1993, schwab_1992, ggq1, ggq2, ggq3, helsing_2008a,
helsing_integral_2009, helsing_tutorial_2012, klockner2013jcp, qbx2,
wala2019jcp, af2018sisc, siegel2018jcp, rachh2017jcp, ding2019arxiv,
bar2014}. For two-dimensional problems, the trapezoid rule is the
workhorse for BIEs since it achieves spectral accuracy when integrands
are not nearly-singular~\cite{tre-wei2014}. To address nearly-singular
integrands of two dimensional BIEs, we will use a {\em barycentric
quadrature rule} that only requires a slight modification of the
trapezoid rule~\cite{ioa-pap-per1991}. In its current form, this method
requires the layer potential satisfy Laplace or Stokes
equations~\cite{bar-wu-vee2015, chi-moo-qua2020}. The PIs will extend
this quadrature rule so that it can be applied to the screened Laplace
equation. This will be done by recognizing that the fundamental solution
of the screened Laplace equation can be decomposed as $K(x,y) = K_1(x,y)
+ K_2(x,y)$, where $K_1(x,y) = -\log\|x - y\|$ and $K_2(x,y) = K(x,y) +
\log\|x - y\|$. Using this decomposition, the layer potential involving
$K_1(x,y)$ can be accurately computed with the barycentric quadrature
rule~\cite{ioa-pap-per1991}, and the layer potential $K_2(x,y)$ can be
accurately computed with the trapezoid rule since the kernel is bounded
for all $x$ and $y$.
\todo[inline]{Image here}

% ----------------------------------------------------------------------
\subsubsection{Avoiding contact with adaptive time stepping and
repulsion}
\label{subsec:timeStepping}
\todo[inline]{Put contact in here as well}
\todo[inline]{Adaptive high-order time stepping}

Even with accurate quadrature and time stepping methods, rigid body
simulations can still result in unphysical contact. We proposed to use
two techniques to avoid unphysical contact: high-order adaptive time
stepping and a repulsion force. PI Quaife developed a high-order
adaptive time stepping method for hydrodynamic
suspensions~\cite{qua-bir2016} and he has used it with colleagues to
investigate mixing and adhesion in suspensions~\cite{qua-vee-you2019,
kab-qua-bir2017}.

In our previous work a
Leonard-Jones potential is used to keep the particles from
overlapping~\cite{Fu2018_SIAM}. Such steep steric interaction at short
ranges introduces great numerical stiffness that limit the time step
size. 
\cite{bys-sha-qua2020}

\textcolor{red}{
Another outstanding numerical issue pertaining to simulating the
self-assembly of amphiphilic particles (such as lipid macromolecles) in
a viscous solvent is the collision between amphiphilic particles. The
hydrophobic attraction potential drives the amphiphilic particles to
move towards each other so to minimize exposure to the solvent. Under
such attraction force the fluid is being squeezed out as particles move
toward each other, and the amphiphilic particles can come into physical
contact in finite time. Such particle collisions in a dense rigid body
suspension is a great challenge and can be a bottleneck in large-scale
simulations. In our previous work a Leonard-Jones potential is used to
keep the particles from overlapping~\cite{Fu2018_SIAM}. Such steep
steric interaction at short ranges introduces great numerical stiffness
that limit the time step size. 
}

\textcolor{red}{
Yan {\it et al.} proposed a collision-resolution algorithm within the
framework of boundary integral methods to resolve the non-smooth
many-body dynamics due to collisions by formulating the particle
collision process as a linear complementarity problem with geometric
`non-overlapping' constraint~\cite{Yan2019}. We propose to treat the
particle collisions based on Yan {\it et al.}'s work after the first two
goals (see above) are achieved. A potential complication in combining
the algorithms for particle collision into our QBX integral formulation
is the computational cost for the QBX to refine meshes on particles in
collision.
}

\textcolor{red}{
Once the collision-resolution algorithm is implemented, we will check
the computing performance first before we embark on incorporating
fluctuating hydrodynamics~\cite{Bao17,Bao18}, which might be important
even at the scales of the amphiphilic particle size in some cases. We
propose to extend the high-order time-integration scheme for a
stochastic differential equation~\cite{fu2015pre} to this system.
}

% ----------------------------------------------------------------------
\subsubsection{Fluctuating hydrodynamics}
\label{subsec:fluctuating}
Using DLP formulation for fluctuating hydrodynamics so that we have a
second-kind BIE

%The HAP model requires solving the exterior Dirichlet problem of the
%screened Laplace equation and the mobility problem of the Stokes flow
%at each time step. Standard numerical methods such as finite difference
%and finite element methods have to truncate the infinite domain to a
%finite computational domain by imposing certain artificial boundary
%conditions on the truncated domain and need the discretization of the
%whole volume of the truncated domain. The artificial boundary
%conditions are very often inexact, introducing another layer of
%approximation that lowers the accuracy of the solution. The
%discretization of the whole volume is very expensive for three
%dimensional problems, leading to excessively large number of unknowns
%in the linear system. Furthermore, it is rather difficult to obtain
%high-order discretization when the boundary surfaces are irregular.

%Potential theory and boundary integral equation (BIE) methods remove the
%aforementioned obstacles in a very elegant way. The starting point is
%that for constant-coefficient PDEs the fundamental solutions, i.e., the
%Green's functions, are readily available. This allows us to represent
%the solution via the so-called layer potentials that are the convolution
%integrals of a kernel and an unknown density only on the boundary. The
%kernel is a linear combination of the Green's function and its
%derivatives, which ensures that the representation satifies the
%underlying PDE and the condition at infinity automatically. And the
%boundary condition at the material interfaces together with the jump
%relation of the layer potential leads to a boundary integral equation
%for the unknown density.

%The BIE method removes the need of imposing artificial boundary
%conditions for exterior problems and reduces the dimension of the
%problem by one, leading to optimal number of unknowns in the solve
%phase. Very often, one is able to construct the so-called {\it second
%kind integral equation} (SKIE) formulation for these problems, where
%the operator is a sum of the identity operator and a compact operator.
%It is well-known that the condition number of the resulting linear
%system for SKIEs is {\it independent} of the mesh size $h$ or
%equivalently, the number of discretization points. This leads to a
%constant number of iterations when iterative solvers are applied to
%solve the resulting linear system. It further improves the accuracy of
%the solution as the relative error of solution will never increase as
%the number of total discretization points increases. On the other hand,
%the condition number of the linear system from finite difference and
%finite element methods generally increases as a power function of the
%total number of dicretization points, with the exact power depending on
%the underlying spatial dimension and the highest order of the
%differential operator in the PDE. This leads to an increase in the
%number of iterations (and hence the total computational cost) and an
%increase in the relative error of the solution as the number of
%dicretization points increases.

%The discretization of the integral operator often leads to a dense
%matrix, while finite difference/finite element methods result in sparse
%matrices, making matrix-vector product much cheaper to compute.
%However,since the invention of the original fast multipole method
%(FMM)\cite{fmm5}, there have been many fast
%algorithms~\cite{fmm1,fmm2,fmm3,fmm4,fmm6,fmm7,fmm8} that reduce the
%cost of matrix-vector product to linear ($O(N)$) or quasilinear
%complexity ($O(N \log^kN)$) for dense matrices resulting from the BIE
%discretization. Fast direct solvers have also been developed
%recently~\cite{fds1,fds2,fds3,fds4,fds5,fds6,fds7,fds8,ho2016cpam2,ho2016cpam1,minden2016,minden2017siammms}.
%The development of these fast algorithms have removed a huge hurdle for
%the use of boundary integral equations, making the BIE method the
%method of choice for constant-coefficient PDEs with linear boundary
%conditions.  For the exterior Dirichlet problem of the screened Laplace
%equation, the SKIE formulation in our previous work~\cite{Fu2018_SIAM}
%can be readily extened to the 3D case. For the mobility problem of the
%Stokes flow in three dimensions that takes the hydrodynamic effect into
%account, an SKIE formulation can be found in~\cite{manasthesis}.

%% ----------------------------------------------------------------------
%\subsubsection{High-order discretization of surface integrals in three
%dimensions}
%The practical application of integral equation methods requires the
%accurate evaluation of boundary integrals with singular, weakly singular
%or nearly singular kernels. There are many numerical quadrature schemes
%for dealing with line integrals in two
%dimensions~\cite{alpert,kapur,sidi,duffy,bruno1,bruno2,davis_1984,graglia_2008,hackbusch_sauter_1994,
%jarvenpaa_2003,khayat_2005,kress_boundary_1991,schwab_1992,
%ying_2006,beale1,beale2,goodman_1990, haroldson_1998,
%lowengrub_1993,schwab_1992,ggq1,ggq2,ggq3,helsing_2008a,helsing_integral_2009,helsing_tutorial_2012}
%and some of them are extremely efficient and can achieve arbitrary high
%order.
%
%The high-order quadrature rules for the evaluation of surface integrals
%in three dimensions, however, are much less developed than the line
%integrals in two dimensions. For example, there are no Gaussian
%quadratures for integraing polynomials on a flat triangle, even though
%efficient quadratures~\cite{xiao2010cma,vioreanu2014} have been
%developed recently for such purpose. For weakly singular or singular
%integrals,~\cite{bremer2012jcp,bremer2013jcp} constructed high-order
%quadratures for surface integrals on a general triangle,
%while~\cite{gimbutas2013sisc} presented a fast algorithm for integrating
%$1/r$-type singular integrals for surfaces that are homeomorphic to a
%sphere. We plan to study the so-called quadrature by expansion (QBX)
%scheme~\cite{klockner2013jcp,qbx2} for the evaluation of both singular
%and near-singular surface integrals encountered in the discretization of
%BIEs in three dimensions. Conceptually, the idea of the QBX to evaluate
%singular, hypersingular and near singular integrals on smooth surfaces
%is straightforward. That is, the surface is discretized into smooth
%triangles and smooth high-order quadratures are applied to evaluate the
%expansion coefficients on all source triangles with the QBX expansion
%center placed at a point off the surface. One may then form a suitable
%expansion (for example, a Taylor expansion) around that center and
%evaluate this expansion back at the target point on the surface (or
%close to the surface in the near singular case). Compared to the
%competing aforementioned quadrature schemes, the QBX quadrature is
%attractive because it offers a clear path for being extended to:
%\textbf{(1)} handle any singularity, including hypersingular operators,
%\textbf{(2)} be usable with any high-order surface discretization,
%\textbf{(3)} generate well-conditioned discrete operators to which
%iterative methods such as GMRES~\cite{gmres} can be applied in a
%black-box fashion, \textbf{(4)} integrate well with fast algorithms such
%as the FMM. 
%
%In practice, there are still many issues that need to be resolved. For
%example, there are now many variants of QBX including global and local
%QBX~\cite{klockner2013jcp,rachh2017jcp}, the target-specific
%QBX~\cite{siegel2018jcp}, kernel-independent QBX~\cite{abtin2018bit},
%and quadrature by two expansions~\cite{ding2019arxiv}. The coupling of
%the QBX and the FMM may also lead to certain instability issues which
%may require some changes in the fast multipole
%method~\cite{wala2018jcp}. Similar to other quadrature methods, there
%have been extensive study on the QBX methods in two dimensions, while
%its three dimension
%version~\cite{wala2019jcp,af2018sisc,siegel2018jcp,wala2019arxiv} has
%not been fully studied and the implementation is even more scarce. We
%plan to investigate the accuracy and the convergence order of the
%various QBX schemes mentioned above, its coupling with the FMM, parallel
%implementation issues for large-scale problems, and the application to
%our target problems.

%% ----------------------------------------------------------------------
%\subsubsection{A hybrid method for close-to-touching Janus particles}
%\textcolor{red}{
%The HAP model for membranes will inevitably lead to the case where some
%Janus particles will be very close to each other. Even though QBX can
%handle such case accurately, it may lead to excessively large number of
%unknowns in order to resolve the unknown densities at those
%close-to-touching points. In~\cite{gan2016sisc}, a hybrid scheme is
%developed to treat the close-to-touching spheres for the interface
%problem of the Laplace equation. The key idea is to combine the analytic
%image method and spectrally accurate method of moments, where the
%analytic image method will capture the most singular part of the unknown
%density at the close-to-touching points. With the aid of the FMM, the
%scheme achieves linear complexity with much less number of unknowns on
%the material interfaces. We plan to study the hybrid scheme to treat
%the close-to-touching case for both the screened Laplace equation and
%the mobility problem to further improve the efficiency of the numerical
%algorithm for the HAP model.
%}
%
%
%% ----------------------------------------------------------------------
%\subsubsection{Other numerical issues}
%Another outstanding numerical issue pertaining to simulating the
%self-assembly of amphiphilic particles (such as lipid macromolecles) in
%a viscous solvent is the collision between amphiphilic particles. The
%hydrophobic attraction potential drives the amphiphilic particles to
%move towards each other so to minimize exposure to the solvent. Under
%such attraction force the fluid is being squeezed out as particles move
%toward each other, and the amphiphilic particles can come into physical
%contact in finite time. Such particle collisions in a dense rigid body
%suspension is a great challenge and can be a bottleneck in large-scale
%simulations. In our previous work a Leonard-Jones potential is used to
%keep the particles from overlapping~\cite{Fu2018_SIAM}. Such steep
%steric interaction at short ranges introduces great numerical stiffness
%that limit the time step size. 
%
%Yan {\it et al.} proposed a collision-resolution algorithm within the
%framework of boundary integral methods to resolve the non-smooth
%many-body dynamics due to collisions by formulating the particle
%collision process as a linear complementarity problem with geometric
%`non-overlapping' constraint~\cite{Yan2019}. We propose to treat the
%particle collisions based on Yan {\it et al.}'s work after the first two
%goals (see above) are achieved. A potential complication in combining
%the algorithms for particle collision into our QBX integral formulation
%is the computational cost for the QBX to refine meshes on particles in
%collision.
%
%Once the collision-resolution algorithm is implemented, we will check
%the computing performance first before we embark on incorporating
%fluctuating hydrodynamics~\cite{Bao17,Bao18}, which might be important
%even at the scales of the amphiphilic particle size in some cases. We
%propose to extend the high-order time-integration scheme for a
%stochastic differential equation~\cite{fu2015pre} to this system.

>>>>

The most significant challenges of a BIE formulation include solving
dense linear systems, developing preconditioning strategies, and
developing quadrature methods for integrands that are nearly singular.
These challenges are described in section~\ref{subsec:NumericalIssues}.
Even with accurate quadrature and time stepping methods, rigid body
simulations can easily result in unphysical contact. We propose two
techniques to avoid unphysical contact in
Section~\ref{subsec:timeStepping}. Then, proposed work for well-conditioned BIE
formulations of fluctuating hydrodynamics are in
section~\ref{subsec:fluctuating}.

@@

Discretizations of carefully chosen BIEs can be solved with a
mesh-independent number of GMRES
iterations~\cite{cam-ips-kel-mey-xue1996}. Therefore, the required CPU
time is proportional to the cost of a matrix-vector multiplication that
can be done in optimal or near-optimal time with the fast multiple
method (FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2, fmm3,
fmm4, fmm6, fmm7, fmm8}. PI Quaife is experienced with applying FMMs for
both the Stokes equation~\cite{qua-bir2014, bys-sha-qua2020} and the
screened Laplace equation~\cite{kro-qua2011, qua2011}. Alternatively,
fast direct solvers can be used to avoid any iteration~\cite{fds1, fds2,
fds3, fds4, fds5, fds6, fds7, fds8, ho2016cpam2, ho2016cpam1,
minden2016, minden2017siammms}, but direct solvers often require a large
amount of initial overhead, and this makes them less useful for problems
with moving geometries. Alternatively, techniques in fast direct solvers
can be used to develop efficient preconditioning strategies such as the
inverse Fast Multipole Method (IFMM)~\cite{cou-pou-dar2017}. PI Quaife
used the IFMM to precondition Stokes equations in a porous
media~\cite{qua-cou-dar2018}, and a suite of other preconditioners
including sparse approximate inverses~\cite{che2000} and
multigrid~\cite{hem-sch1981, sch1982} will be investigated.

Nearly touching bodies is ubiquitous in self-assembly of amphiphilic
particles, and such near-contact results in nearly-singular integrands
in both two and three dimensions. Quadrature methods to address such
integrands has received a lot of attention in two and three
dimensions~\cite{alpert, kapur, sidi, duffy, bruno1, bruno2, davis_1984,
graglia_2008, hackbusch_sauter_1994, jarvenpaa_2003, khayat_2005,
schwab_1992, ying_2006, beale1, beale2, goodman_1990, haroldson_1998,
lowengrub_1993, schwab_1992, ggq1, ggq2, ggq3, helsing_2008a,
helsing_integral_2009, helsing_tutorial_2012, klockner2013jcp, qbx2,
wala2019jcp, af2018sisc, siegel2018jcp, rachh2017jcp, ding2019arxiv,
bar2014}. For two-dimensional problems, the trapezoid rule is the
workhorse for BIEs since it achieves spectral accuracy when integrands
are not nearly-singular~\cite{tre-wei2014}. To address nearly-singular
integrands of two dimensional BIEs, we will use a {\em barycentric
quadrature rule} that only requires a slight modification of the
trapezoid rule~\cite{ioa-pap-per1991}. In its current form, this method
requires the layer potential satisfy Laplace or Stokes
equations~\cite{bar-wu-vee2015, chi-moo-qua2020}. The PIs will extend
this quadrature rule so that it can be applied to the screened Laplace
equation. This will be done by recognizing that the fundamental solution
of the screened Laplace equation can be decomposed as $K(x,y) = K_1(x,y)
+ K_2(x,y)$, where $K_1(x,y) = -\log\|x - y\|$ and $K_2(x,y) = K(x,y) +
\log\|x - y\|$. Using this decomposition, the layer potential involving
$K_1(x,y)$ can be accurately computed with the barycentric quadrature
rule~\cite{ioa-pap-per1991}, and the layer potential $K_2(x,y)$ can be
accurately computed with the trapezoid rule since the kernel is bounded
for all $x$ and $y$.

\includegraphics[scale=1]{figures/integrands.pdf}

% ----------------------------------------------------------------------
\subsubsection{Avoiding contact with adaptive time stepping and
repulsion}
\label{subsec:timeStepping}

An outstanding numerical issue when simulating the
self-assembly of amphiphilic particles in a viscous solvent is the
collision between amphiphilic particles. The hydrophobic attraction
potential drives the amphiphilic particles towards one another so to
minimize exposure to the solvent, and this leads to physical contact in
finite time. Such particle collisions in a dense rigid body suspension
is a great challenge and can be a bottleneck in large-scale simulations.
We propose two algorithms to remedy this computational
challenge---high-order adaptive time stepping and repulsion forces.

PI Quaife developed a high-order adaptive time stepping method for
hydrodynamic suspensions~\cite{qua-bir2016} and he has used it with
colleagues to investigate mixing and adhesion in
suspensions~\cite{qua-vee-you2019, kab-qua-bir2017}. The method requires
both a single-step high-order time stepping method and a computationally
cheap estimate of the error. We plan to use a spectral deferred
correction method~\cite{dut-gre-rok2000} since it can be used
iteratively to increase the order of accuracy. To estimate the error,
instead of using an embedded Runge-Kutta method or step-doubling, we
will compute the total force and torque of the system which is
computationally cheap and should be zero. The PIs experience is that
error estimates based on physical constraints such as force- and
torque-free conditions appropriately adjusts the time step size so that
the dynamics are adequately resolved.

Our previous work avoids contact by including a Leonard-Jones potential
in the body forces~\cite{Fu2018_SIAM}. Such steep steric interaction at
short ranges introduces great numerical stiffness that limit the time
step size. To remove this stiffness, we will use a geometry-based
contact method~\cite{har-pon-sor-zor2011}, and this method has been
applied to vesicle suspensions~\cite{lu-rah-zor2017} and rigid body
suspensions in two dimensions~\cite{bys-sha-qua2020} and
three dimensions~\cite{Yan2019}. These methods determine repulsion
force by solving a non-linear complementarity problem with a geometric
constraint that the configuration is non-overlapping.

% ----------------------------------------------------------------------
\subsubsection{Fluctuating hydrodynamics}
\label{subsec:fluctuating}
Using DLP formulation for fluctuating hydrodynamics so that we have a
second-kind BIE
\textcolor{red}{
Once the collision-resolution algorithm is implemented, we will check
the computing performance first before we embark on incorporating
fluctuating hydrodynamics~\cite{Bao17,Bao18}, which might be important
even at the scales of the amphiphilic particle size in some cases. We
propose to extend the high-order time-integration scheme for a
stochastic differential equation~\cite{fu2015pre} to this system.
}
